---
title: "Brain Somatic variants and aging"
subtitle: 'PsychAD aging data'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
output: 
  html_document:
    toc: true
    smart: true
    self_contained: true
---

<!---

system("rm -rf aging_testing_cache/")
rmarkdown::render("aging_testing.Rmd");


--->


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```


```{r load, cache=FALSE}
library(glmmTMB)
library(lme4)
library(tidyverse)
library(ggplot2)
library(parallel)
library(kableExtra)
library(lmerTest)

control_data = readRDS("~/Downloads/Aging_data_all_CellType.rds")
```

### Somatic variants versus callable sites
```{r Callable_Sites}
ggplot(control_data, aes(Callable_Sites, Count+.5, color=log(no_reads))) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1) +
  facet_wrap(~CellType)

ggplot(control_data, aes(CellType, Callable_Sites, fill=CellType)) +
  geom_violin() +  
  geom_boxplot(width=.1) +
  scale_y_log10() +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1, legend.position = "none")
```

### Fit regression models for each cell type
```{r regression}
ctrl = glmmTMBControl(optimizer=optim, optArgs=list(method="BFGS"))

resList = mclapply( unique(control_data$CellType), function(CT){

  # Filter data to only Donors with replicates
  keep = control_data$CellType == CT
  data = control_data[keep,]
  tab = table(data$SubID.x)
  include = names(tab)[tab > 1]
  data = data[data$SubID.x %in% include,]
  
  # baseline formula
  form = Count ~ offset(log(Callable_Sites)) + (1|SubID.x) + log(no_cells) + log(no_reads)

  # fit NB mixed model for baseline model
  fit0 <- glmmTMB(form, family=nbinom2, data)

  # if model didn't converge, try again with different
  # optimizer
  if( fit0$fit$convergence != 0){
    fit0 = update(fit0, control=ctrl)
  }

  # fit null model, then linear and quadratic
  fitNull <- update(fit0, ~ . + PMI  + Sex + Brain_bank + (1|Ethnicity))
  fitLinear = update(fitNull, ~ . + Age)
  fitQuad = update(fitLinear, ~ . + I(Age^2))

  fitQuad = update(fitQuad, control=ctrl)

  list(CellType = CT, 
        data = data,
        fit0 = fit0, 
        fitNull = fitNull, 
        fitLinear = fitLinear, 
        fitQuad = fitQuad)
}, mc.cores=6)
```

`form = Count ~ offset(log(Callable_Sites)) + (1|SubID.x) + log(no_cells) + log(no_reads) + PMI  + Sex + Brain_bank + (1|Ethnicity) + VARIABLE`



# Hypothesis testing
```{r tests}
# Test linear Age term for each model
resLinear = lapply(resList, function(x){
  df = coef(summary(x$fitLinear))$cond["Age",]
  data.frame(CellType = x$CellType, t(df))
  })
resLinear = do.call(rbind, resLinear) %>%
            rename(p.value = Pr...z..) %>%
            mutate(FDR = p.adjust(p.value, "fdr"))

resLinear %>%
  mutate(CellType = factor(CellType, CellType)) %>%
  ggplot(aes(CellType, Estimate, color=CellType)) +
    geom_point() +
    geom_errorbar(aes(ymin = Estimate - 1.96*Std..Error, ymax = Estimate + 1.96*Std..Error), width=0) +
    geom_text(aes(CellType, .03, label=format(FDR, digits=3)), color="black") +
    theme_classic() +
    theme(aspect.ratio=1, legend.position="none") +
    geom_hline(yintercept=0, linetype="dashed") +
    coord_flip() +
    ggtitle("Linear")

resLinear %>%
  mutate( p.value = format(p.value, scientific=TRUE)) %>%
  kbl %>% 
  kable_classic(full_width = FALSE)

# Test quadratic Age term for each model
resQuad = lapply(resList, function(x){
  df = coef(summary(x$fitQuad))$cond["I(Age^2)",]
  data.frame(CellType = x$CellType, t(df))
  })
resQuad = do.call(rbind, resQuad) %>%
            rename(p.value = Pr...z..) %>%
            mutate(FDR = p.adjust(p.value, "fdr"))

resQuad %>%
  mutate(CellType = factor(CellType, CellType)) %>%
  ggplot(aes(CellType, Estimate, color=CellType)) +
    geom_point() +
    geom_errorbar(aes(ymin = Estimate - 1.96*Std..Error, ymax = Estimate + 1.96*Std..Error), width=0) +
    geom_text(aes(CellType, .001, label=format(FDR, digits=3)), color="black") +
    theme_classic() +
    theme(aspect.ratio=1, legend.position="none") +
    geom_hline(yintercept=0, linetype="dashed") +
    coord_flip() +
    ggtitle("Quadratic")

resQuad %>%
  mutate( p.value = format(p.value, scientific=TRUE)) %>%
  kbl %>% 
  kable_classic(full_width = FALSE)
```

### Estimate somatic rate per donor
Somatic rate is the number of somatic variants per 1M callable sites, 1M reads and 1K cells.

```{r ranef}
df = lapply(resList, function(x){

  # predictions at donor level, centered at zero
  ran = ranef(x$fit0, condVar=TRUE)
  se = attr( ran$cond$SubID.x, "condVar")
  SomaticRate = ran$cond$SubID.x$`(Intercept)`

  # baseline somatic rate based on cells, reads and callable sites
  no_cells = 1000
  no_reads = 1000000
  Callable_Sites = 1000000
  baseline = fixef(x$fit0)$cond %*% c(intercept = 1,log(no_cells),log(no_reads)) + log(Callable_Sites)
  baseline = c(baseline)

  SomaticRate = SomaticRate + baseline

  i = match(rownames(ran$cond$SubID.x), control_data$SubID.x)

  data.frame(
        CellType = x$CellType, 
        SubId = rownames(ran$cond$SubID.x),
        Age = control_data$Age[i],
        SomaticRate = SomaticRate,
        se = c(attr( ran$cond$SubID.x, "condVar")))
  })
df = do.call(rbind, df)
df = df %>% 
  droplevels %>%
  mutate(CellType = factor(CellType, sort(levels(CellType))))
```


```{r plots.ranef}
df %>% 
  ggplot(aes(Age, SomaticRate)) +
  geom_point(aes(color=se)) +
  theme_classic() +
  geom_smooth(method = "lm", se=FALSE) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), color="green3", se=FALSE) +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1) +
  facet_wrap(~CellType) +
  scale_color_gradient(low="grey", high="red") +
  ylab("Estimated somatic rate")
```

#### Including standard error of the somatic rate 
```{r plots.ranef.weights}
df %>% 
  ggplot(aes(Age, SomaticRate, weight=1/se^2)) +
  geom_point(aes(color=se)) +
  theme_classic() +
  geom_smooth(method = "lm", se=FALSE) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), color="green3", se=FALSE) +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1) +
  facet_wrap(~CellType) +
  scale_color_gradient(low="grey", high="red") +
  ylab("Estimated somatic rate")
```

```{r plots.ranef.weights.scale, fig.width=10, fig.height=5}
df %>% 
  ggplot(aes(Age, SomaticRate, weight=1/se^2, size=1/se^2)) +
  geom_point(aes(color=se)) +
  theme_classic() +
  geom_smooth(method = "lm", se=FALSE) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), color="green3", se=FALSE) +
  theme(plot.title = element_text(hjust = 0.5), aspect.ratio=1) +
  facet_wrap(~CellType, scale="free_y") +
  scale_color_gradient(low="grey", high="red") +
  ylab("Estimated somatic rate") +
  scale_size(range=c(.1, 4))
```


### Approximate results with two-step model
```{r test.of.ranef}
resLM = lapply(levels(df$CellType), function(CT){

  df2 = df %>%
    filter(CellType == CT)

  fit = lm(SomaticRate ~ Age , df2, weight = 1/se^2)
  data.frame(CellType = CT, t(coef(summary(fit))[2,]))
  })
resLM = do.call(rbind, resLM)

resLM %>%  
  rename(p.value = Pr...t..) %>%
  mutate( p.value = format(p.value, scientific=TRUE)) %>%
  kbl %>% 
  kable_classic(full_width = FALSE)
```







